--- 
title: "Notes for Grad Student Orientation 2018"
author: "Charles J. Geyer"
site: bookdown::bookdown_site
output: 
  bookdown::gitbook:
    config:
      toc:
        collapse: section
        depth: 2
        scroll_highlight: yes
        before: null
        after: null
      toolbar:
        position: fixed
      edit : null
      download: null
      search: yes
      fontsettings:
        theme: white
        family: sans
        size: 2
      sharing:
        facebook: no
        twitter: no
        google: no
        weibo: no
        instapper: no
        vk: no  
  bookdown::pdf_book: default
  bookdown::epub_book: default
documentclass: book
github-repo: cjgeyer/Orientation2018/docs
---

```{r set-options, include=FALSE}
options(width = 72, digits = 4)
knitr::opts_chunk$set(tidy = FALSE, out.width='\\textwidth',
                      fig.align = "center",
                      fig.width=3.5,
                      fig.height=2.75,
                      collapse=TRUE,
                      rows.print = 25,
                      warning=FALSE)
```

# Introduction

These are some notes for grad student orientation in the School of Statistics,
University of Minnesota, Fall 2018.

## License

This work is licensed under a Creative Commons
Attribution-ShareAlike 4.0 International License
(<http://creativecommons.org/licenses/by-sa/4.0/>).


## R

 * The version of R used to make this document is `r getRversion()`.

 * The version of the `bookdown` package used to make this document is `r packageVersion("bookdown")`.

 * The version of the `rmarkdown` package used to make this document is `r packageVersion("rmarkdown")`.

 * The version of the `knitr` package used to make this document is `r packageVersion("knitr")`.

 * The version of the `ggplot2` package used to make this document is `r packageVersion("ggplot2")`.

## Other Learning Materials

### An Introduction to R

By far the best book on R is free, written by the R core team, always up to
date with the current version, and always correct.  It is called
*An Introduction to R* and can be found in your R distribution.  Do
```{r eval=FALSE}
help.start()
```
to start browser help and click on the "An Introduction to R" link.

If can also be found [on-line at CRAN](https://cloud.r-project.org/doc/manuals/r-release/R-intro.html).

[PDF and e-book versions are also available at CRAN](https://cloud.r-project.org/manuals.html).

On the web page `help.start()` gives you or
at the link just above you see there are also
five other manuals that come with R, but they are for experts.
You don't want to read them yet.

### An R Short Course

Your humble author was one of five instructors for a two-day short course
on R.  [Here are the notes for it](https://irsaatumn.github.io/RWorkshop18/).

### An R Course (Undergraduate)

Your humble author taught Stat 3701 (undergraduate statistical computing).
[Here is the web site for that](http://www.stat.umn.edu/geyer/3701/).

Of particular interest are the

 * [reproducible research examples](http://www.stat.umn.edu/geyer/3701/repro/).

 * [course notes](http://www.stat.umn.edu/geyer/3701/notes/).

### An R Course (PhD Level)

Your humble author taught Stat 8054 (PhD level statistical computing).
[Here is the web site for that](http://www.stat.umn.edu/geyer/8054/).

That site is a bit out of date.  I will redo it when I teach that
course again in Spring 2019, including many topics from my Stat 3701 notes
linked above, like web scraping, JSON, and SQL databases.
And I will update many topics, like parallel computing.
